---
title: "Extracting compositions from Seurat and anndata"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    md_extensions: +tex_math_single_backslash
    pandoc_args: ["--mathjax"]
vignette: >
  %\VignetteIndexEntry{Extracting compositions from Seurat and anndata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Single-cell pipelines store the annotations that define a composition at the cell level, while Hellinger enrichment operates on subject-level category counts. The gap between those scales is not a modeling choice; it is a bookkeeping step that must be correct before any distance or permutation test runs. Expression assays, normalized layers, and gene-level matrices are irrelevant to this step. What matters is whether each cell can be assigned to a subject, a categorical label (cluster, cell type, or other partition), and an experimental group that is constant for that subject.

Here, we show how `ExtractClusterComposition()` bridges common single-cell containers to the `CategoryComposition` object used by `CompareGroupCompositions()`. The companion vignette on Hellinger enrichment covers the distance metric, enrichment ratio, and inference methods; this article focuses on getting the count table right.

## Concepts

### What metadata does enrichment need?

Composition enrichment asks whether experimental groups differ in how subjects distribute cells across categories. That question is defined entirely by three metadata fields per cell: which subject the cell came from, which category it belongs to, and which group that subject is in. The mapping from cells to those three columns is what downstream analysis consumes; everything else in a Seurat object or AnnData file is carried along but not read.

At the subject level, the analysis needs a non-negative count matrix. Each row is one subject, each column is one category, and entry \(c_{sk}\) is the number of cells (or pre-aggregated count) assigned to category \(k\) for subject \(s\). If a subject has no cells in a category, the corresponding entry is zero rather than missing. Group membership is stored once per subject and must not vary across cells from the same subject, because enrichment compares between-group distances to within-group distances among subjects, not among cells.

### Why bridge cell-level objects to subject-level counts?

The scale transfer is deliberate. A typical single-cell dataset contains thousands or millions of cells but only a handful of biological replicates. Treating each cell as an independent observation would inflate replication and distort the null distribution of group differences. Aggregating to subjects collapses the cell-level multinomial into one frequency vector per donor, which is the unit at which group comparisons are defined.

The container format does not change that aggregation. The extractor accepts Seurat `meta.data`, AnnData `obs`, or a plain `data.frame`, tabulates cells by subject and category (or sums a count column in a long table), and returns the same `CategoryComposition` structure. Once that object exists, the enrichment workflow is identical regardless of how the counts were produced. Object type is therefore an interface concern, not a statistical one.

### How are columns named and validated?

Column names in real pipelines rarely match package defaults, so `ExtractClusterComposition()` accepts explicit mappings: `subjectCol` for the donor or sample identifier, `categoryCol` for the cluster or cell-type label, and `groupCol` for the experimental condition. For pre-aggregated long tables, a fourth column `countCol` (default `"n"`) holds the raw counts. The function checks that required columns exist, coerces identifiers to character, and stops if any subject maps to more than one group.

Counts must be non-negative, and the composition object does not permit `NA` in the count matrix; absent subject–category combinations are represented as zero. Inference via `CompareGroupCompositions()` additionally requires at least two subjects per group so that within-group distances can be computed. These constraints are enforced at extraction or at comparison time depending on the failure mode, but the design intent is the same: subject-level replication with a single, unambiguous group label per subject.

## Execution

### Extract compositions from Seurat `meta.data`

For Seurat objects, `ExtractClusterComposition()` reads the requested columns from `meta.data`, treats each row as one cell, and aggregates to subject-level counts. We build a minimal in-memory object so the extraction can run without a study-sized dataset; the expression matrix is present only because Seurat requires one, not because enrichment uses it.

After extraction, `print(composition)` reports the number of subjects, categories, and groups. A short `CompareGroupCompositions()` call with reduced permutations confirms that the returned object is a valid input to the enrichment workflow described in the companion vignette.

```{r seurat, eval=requireNamespace("Seurat", quietly = TRUE)}
library(HellingerDistanceEnrichment)

if (requireNamespace("Seurat", quietly = TRUE)) {
  n_cells <- 120
  meta <- data.frame(
    subjectId = rep(paste0("Donor", 1:6), each = 20),
    category = sample(paste0("Cluster", 0:3), n_cells, replace = TRUE),
    group = rep(c("Control", "Control", "Control", "Case", "Case", "Case"), each = 20),
    stringsAsFactors = FALSE
  )

  counts <- matrix(
    stats::rpois(100 * n_cells, lambda = 2),
    nrow = 100,
    ncol = n_cells
  )
  colnames(counts) <- paste0("cell", seq_len(n_cells))
  rownames(counts) <- paste0("gene", seq_len(100))

  seu <- Seurat::CreateSeuratObject(counts = counts, meta.data = meta)

  composition <- ExtractClusterComposition(
    seu,
    subjectCol = "subjectId",
    categoryCol = "category",
    groupCol = "group"
  )

  print(composition)

  result <- CompareGroupCompositions(
    composition,
    method = "permutation",
    nPermutations = 50,
    seed = 1
  )
  result$omnibus
} else {
  message("Seurat not installed; skipping Seurat example.")
}
```

### Extract compositions from anndata `obs`

The AnnData approach is mechanically the same: cell-level fields live in `obs`, and the extractor delegates to the same `data.frame` aggregation after coercing `obs` to a table. Column names may differ from the Seurat example (`subjectId` versus `donor_id`, `category` versus `cell_type`), which is why the `subjectCol`, `categoryCol`, and `groupCol` arguments exist. Once those names are aligned with the object at hand, the resulting `CategoryComposition` is interchangeable with one derived from Seurat.

```{r anndata, eval=requireNamespace("anndata", quietly = TRUE)}
if (requireNamespace("anndata", quietly = TRUE)) {
  obs <- data.frame(
    subjectId = rep(paste0("S", 1:4), each = 15),
    category = sample(paste0("Type", 0:2), 60, replace = TRUE),
    group = rep(c("A", "A", "B", "B"), each = 15),
    stringsAsFactors = FALSE
  )
  X <- matrix(stats::rnorm(60 * 10), nrow = 60)
  adata <- anndata::AnnData(X = X, obs = obs)

  composition <- ExtractClusterComposition(
    adata,
    subjectCol = "subjectId",
    categoryCol = "category",
    groupCol = "group"
  )

  CompareGroupCompositions(
    composition,
    nPermutations = 30,
    seed = 2
  )$omnibus
} else {
  message("anndata not installed; skipping anndata example.")
}
```

### Use a long counts table directly

Not every workflow passes through a single-cell object. If counts are already aggregated—one row per subject–category combination with an explicit count column—a long `data.frame` is a first-class input. When all four columns (`subjectCol`, `categoryCol`, `groupCol`, `countCol`) are present, `ExtractClusterComposition()` detects the long-table format and sums duplicate subject–category rows rather than counting cells. The same table can be passed directly to `CompareGroupCompositions()` without calling the extractor first, though extracting explicitly makes the subject×category matrix visible and attaches provenance metadata.

This approach is useful when metadata were exported from a database, pseudobulked upstream, or assembled outside R. The statistical object on the other side is unchanged.

```{r long-table}
long_counts <- data.frame(
  subjectId = rep(c("S1", "S2", "S3", "S4"), each = 3),
  category = rep(paste0("Category", 0:2), 4),
  group = rep(c("Control", "Control", "Case", "Case"), each = 3),
  n = c(30, 30, 40, 35, 25, 40, 10, 20, 70, 15, 15, 70),
  stringsAsFactors = FALSE
)

composition <- ExtractClusterComposition(long_counts)
print(composition)

CompareGroupCompositions(
  long_counts,
  method = "permutation",
  nPermutations = 50,
  seed = 3
)$omnibus
```

### Deploy on HPC or in containers

In production, cohort annotations often change faster than container images or batch job definitions. Embedding subject and group labels in an image layer couples the enrichment code to a specific study snapshot and complicates reruns when metadata are corrected. Providing metadata at runtime—as exported `meta.data`, `obs`, or a long counts CSV—keeps the extraction step portable while the cohort files remain outside the image.

The extraction function depends only on whether the three (or four) columns validate, not on where the table originated. That separation allows the same Docker image or HPC module to serve multiple studies by swapping mounted metadata files, with `subjectCol` / `categoryCol` / `groupCol` adjusted per cohort as needed.

Object format therefore does not dictate the enrichment workflow. Seurat, AnnData, and long tables are alternative inputs to the same `CategoryComposition` object. Once counts are at subject scale with one group per subject, the distance and inference steps in the Hellinger enrichment vignette apply without modification.
