---
title: "Extracting compositions from Seurat and anndata metadata"
author: "HellingerDistanceEnrichment"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting compositions from Seurat and anndata metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Single-cell workflows store the annotations needed for composition analysis at
the cell level, while Hellinger enrichment operates on subject-level category
counts. Bridging those scales depends on metadata columns that map each cell to
a subject, a category (cluster or cell type), and an experimental group. This
vignette builds tiny synthetic Seurat and anndata objects in memory and shows
that both routes produce the same `CategoryComposition` input used by
`CompareGroupCompositions()`.

## Seurat metadata path

For a Seurat object, `ExtractClusterComposition()` reads the requested columns
from `meta.data`, aggregates cells to subjects, and returns the composition
object. Expression counts are irrelevant to the enrichment step; only the
metadata mapping matters. The example below creates a minimal object so the
extraction path can be run without a study-sized dataset.

```{r seurat, eval=requireNamespace("Seurat", quietly = TRUE)}
library(HellingerDistanceEnrichment)

if (requireNamespace("Seurat", quietly = TRUE)) {
  n_cells <- 120
  meta <- data.frame(
    subjectId = rep(paste0("Donor", 1:6), each = 20),
    category = sample(paste0("Cluster", 0:3), n_cells, replace = TRUE),
    group = rep(c("Control", "Control", "Control", "Case", "Case", "Case"), each = 20),
    stringsAsFactors = FALSE
  )

  counts <- matrix(
    stats::rpois(100 * n_cells, lambda = 2),
    nrow = 100,
    ncol = n_cells
  )
  colnames(counts) <- paste0("cell", seq_len(n_cells))
  rownames(counts) <- paste0("gene", seq_len(100))

  seu <- Seurat::CreateSeuratObject(counts = counts, meta.data = meta)

  composition <- ExtractClusterComposition(
    seu,
    subjectCol = "subjectId",
    categoryCol = "category",
    groupCol = "group"
  )

  print(composition)

  result <- CompareGroupCompositions(
    composition,
    method = "permutation",
    nPermutations = 50,
    seed = 1
  )
  result$omnibus
} else {
  message("Seurat not installed; skipping Seurat example.")
}
```

## anndata obs path

The anndata route is mechanically the same: cell-level fields live in `obs`,
and the extractor aggregates them to subjects. Column names may differ across
pipelines, so the `subjectCol`, `categoryCol`, and `groupCol` arguments must
match the object at hand. Once those columns are identified, the resulting
composition object is interchangeable with the Seurat-derived one.

```{r anndata, eval=requireNamespace("anndata", quietly = TRUE)}
if (requireNamespace("anndata", quietly = TRUE)) {
  obs <- data.frame(
    subjectId = rep(paste0("S", 1:4), each = 15),
    category = sample(paste0("Type", 0:2), 60, replace = TRUE),
    group = rep(c("A", "A", "B", "B"), each = 15),
    stringsAsFactors = FALSE
  )
  X <- matrix(stats::rnorm(60 * 10), nrow = 60)
  adata <- anndata::AnnData(X = X, obs = obs)

  composition <- ExtractClusterComposition(
    adata,
    subjectCol = "subjectId",
    categoryCol = "category",
    groupCol = "group"
  )

  CompareGroupCompositions(
    composition,
    nPermutations = 30,
    seed = 2
  )$omnibus
} else {
  message("anndata not installed; skipping anndata example.")
}
```

The practical consequence is that object format need not dictate the enrichment
workflow. Seurat and anndata are alternative containers for the same
subject-category-group mapping. For HPC deployments, bind-mount study metadata
at runtime rather than baking cohort files into container images; the extraction
step then remains portable across environments while the cohort annotations stay
outside the image layer.
