---
title: "Hellinger enrichment for subject-level compositions"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    md_extensions: +tex_math_single_backslash
    pandoc_args: ["--mathjax"]
vignette: >
  %\VignetteIndexEntry{Hellinger enrichment for subject-level compositions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

Understanding whether categorical compositions differ between experimental groups is critical when single-cell annotations are summarized to subject-level frequencies. However, the scale of inference is not the cell: it is the donor (or animal, or replicate) whose multinomial counts define a composition vector on a probability simplex. Cell-level cluster proportions can look sharply different while subject-level compositions remain exchangeable across groups, and the reverse can hold when sampling noise is large relative to an imposed enrichment. Real cohorts rarely reduce to a single pairwise comparison, so this vignette uses three groups: a Control, a Condition1 group with an imposed enrichment, and a Condition2 group that remains exchangeable with Control. Here, we illustrate `HellingerDistanceEnrichment` on that synthetic cohort, separating the conceptual framework (distance, ratio, contrasts, inference) from the executable workflow that recovers which comparisons carry signal.

## Concepts

### What is composition enrichment?

Composition enrichment asks whether experimental groups occupy different regions of the simplex once cells have been aggregated to subjects. At the cell scale, each row carries a category label (cluster, cell type, module score bin) and metadata that links that cell to a subject and a group. At the subject scale, those labels become counts: for subject \(s\) and category \(k\), we observe \(c_{sk}\) cells (or events) in category \(k\), and the subject's composition is the normalized frequency vector \(p_s = (c_{s1}, \ldots, c_{sK}) / \sum_k c_{sk}\).

The scientific question is relational rather than descriptive. We are not asking which category is most abundant in one subject; we are asking whether between-group compositional distance exceeds within-group compositional distance across subjects. That framing mirrors pseudobulking in expression analysis: many cells collapse to one replicate-level observation, and hypothesis tests must respect the replicate as the unit of inference. A cluster that is enriched at the cell level within a single subject does not, by itself, establish group-level enrichment, because one subject's multinomial draw can dominate the apparent pattern.

This package tests enrichment by comparing mean pairwise Hellinger distances among subjects in different groups to mean pairwise distances among subjects in the same group. The output is an effect-size ratio together with permutation or posterior-predictive p-values, optionally decomposed into omnibus and contrast-specific summaries.

### Why Hellinger distance?

Once counts are normalized to frequencies, compositions live on a simplex, and Euclidean distance on raw proportions is a poor geometry: scaling one category compresses the others, and small zeros carry disproportionate weight. Hellinger distance measures separation between two frequency vectors \(p\) and \(q\) by comparing their square roots:

\[
H(p,q)=\frac{1}{\sqrt{2}}\left\lVert\sqrt{p}-\sqrt{q}\right\rVert_2 \in [0,1].
\]

Values near 0 indicate similar compositions; values near 1 indicate near-disjoint support. The implementation computes \(H\) from count vectors by forming frequencies internally, so the distance is always taken on normalized compositions rather than raw counts.

Sparse categories create a second problem: empirical frequencies of zero are not credible endpoints when counts are small. Before distances are evaluated, counts are softened with a Jeffreys-style pseudocount. For subject \(s\), category \(k\), and default \(\alpha = 1/2\),

\[
\tilde{c}_{sk} = c_{sk} + \alpha, \qquad
p_{sk} = \frac{\tilde{c}_{sk}}{\sum_j \tilde{c}_{sj}}.
\]

The default `priorPseudocounts = 1/2` matches the conjugate Dirichlet prior used in the Bayesian approach. Softening pulls extreme proportions toward the simplex center and stabilizes pairwise distances when some categories are rarely observed.

Hellinger distance is appropriate for asking whether compositions are separable in frequency space. It does not, by itself, identify which category drives a contrast, establish causality, or correct for differences in total cell yield between subjects beyond what is encoded in the normalized frequencies. Biological interpretation still requires knowing what the categories represent and whether the aggregation level matches the experimental question.

### What is the enrichment ratio?

Given a subjects-by-categories count matrix, the package builds a symmetric subject-level Hellinger distance matrix \(D\), where \(D_{ss'} = H(p_s, p_{s'})\) using softened, normalized compositions. Subjects are then grouped by experimental label \(g\). For each group pair \((g, g')\), the package summarizes distances as mean pairwise Hellinger distance within or between subject sets:

- **Within-group** (diagonal of the group summary): mean of \(D_{ss'}\) for subjects \(s, s'\) both in group \(g\).
- **Between-group** (off-diagonal): mean of \(D_{ss'}\) for subjects \(s\) in group \(g\), \(s'\) in group \(g'\).

The enrichment effect size is the ratio of average between-group distance to average within-group distance:

\[
R=\frac{\overline{H}_{\mathrm{between}}}{\overline{H}_{\mathrm{within}}}.
\]

When \(R > 1\), subjects in different groups are farther apart, on average, than subjects within the same group under this metric. When \(R \approx 1\), group labels are consistent with exchangeable compositions relative to the observed within-group dispersion. The ratio is defined over pairwise subject distances, so each group must contribute at least two subjects; otherwise within-group means are not identifiable and inference stops with an error.

This ratio is the test statistic for both permutation and Bayes methods. It compresses a full distance matrix into a single scale-free quantity that can be compared across contrasts and simulation draws.

### What is a contrast?

A contrast is a specified comparison among group labels on which \(R\) is evaluated. The package always computes an **omnibus** contrast that includes all group levels present in the composition object. For \(G\) groups, it also computes all **pairwise** contrasts (\(\binom{G}{2}\) comparisons) unless additional custom specifications are supplied.

The analogy to differential expression is deliberate but not exact. In DE, a contrast selects linear combinations of model coefficients on log-expression. Here, a contrast selects which subjects enter the between/within distance summaries—either all groups (omnibus), a pair of groups (subset), or a remapped grouping (collapse). Custom contrasts are passed as a **named list** to `contrasts=`; names become `contrastId` values in the result.

Two levels of p-value reporting matter in practice:

- **`result$omnibus$pValue`**: unadjusted p-value for the global between/within ratio across all groups.
- **`result$contrasts$pValue` and `pAdj`**: contrast-specific p-values, with `pAdj` applying Holm adjustment across the contrast table by default (`pAdjustMethod = "holm"`).

Omnibus and pairwise contrasts therefore answer nested questions. A significant omnibus result indicates that at least some group separation exists under the ratio metric; pairwise contrasts localize that separation, with multiplicity control on the contrast table rather than on the omnibus line.

### Permutation vs Bayes: related but non-equivalent questions

Both methods use label shuffling to define a null, but they differ in whether composition vectors are treated as fixed.

**Permutation (`method = "permutation"`).** Compositions are estimated once from softened counts, fixing the subject distance matrix. Group labels are permuted among subjects \(B\) times (`nPermutations`), and \(R\) is recomputed on each shuffle. The one-sided p-value includes the observed statistic in the numerator:

\[
p=\frac{\#\{R_{\mathrm{perm}}\ge R_{\mathrm{obs}}\}+1}{B+1}.
\]

This tests whether group labels are exchangeable **given** the estimated compositions. It does not propagate uncertainty in the multinomial allocation of cells to categories.

**Bayes (`method = "bayes"`).** For each posterior draw, subject compositions are sampled from a conjugate Dirichlet posterior. With Jeffreys increment \(\alpha\),

\[
p_s \sim \mathrm{Dirichlet}(c_{s\cdot} + \alpha),
\]

where \(c_{s\cdot}\) is the count vector for subject \(s\). A full Hellinger distance matrix is rebuilt from that draw, and a nested permutation null (same label shuffles, `nPermutations` per draw) yields \(R\) and a draw-specific p-value. Across `nPosterior` draws, the package reports:

- posterior-mean \(R\) (`effectSize`),
- posterior-mean p-value (`pValue`),
- 95% credible interval on \(R\) (`effectCiLow`, `effectCiHigh`).

The Bayesian approach therefore nests composition uncertainty inside the label-exchange null. When counts are sparse, that nesting can widen credible intervals or inflate mean p-values relative to plain permutation, even when point ratios look similar. When counts are dense and the imposed enrichment is strong—as in the synthetic example below—the two approaches often agree on which contrasts are elevated, though they still answer formally different questions.

## Execution

### Build a synthetic long counts table

We begin with a known enrichment structure. A synthetic long table carries one row per subject–category combination with columns `subjectId`, `category`, `group`, and `n`, matching the package's default long-table schema. Subjects are drawn from a multinomial allocation within each group so that, before enrichment is imposed, Control, Condition1, and Condition2 are approximately exchangeable. We then multiply counts in `Condition1` for `Category0` by a fixed factor, introducing a modest, known enrichment in that group alone. Condition2 is left unmodified, so its compositions remain exchangeable with Control under the generating process.

The internal functions `build_synthetic_long_table()` and `plant_group_structure()` live in the package namespace for vignette and test use; they are not exported, but they provide a reproducible procedure for simulating compositional cohorts.

```{r synthetic-data}
library(HellingerDistanceEnrichment)

set.seed(2026)
long_table <- HellingerDistanceEnrichment:::build_synthetic_long_table(
  n_subjects_per_group = 6,
  n_categories = 5,
  groups = c("Control", "Condition1", "Condition2"),
  seed = 2026
)

long_table <- HellingerDistanceEnrichment:::plant_group_structure(
  long_table,
  target_group = "Condition1",
  target_category = "Category0",
  boost = 4
)

head(long_table)
table(unique(long_table[, c("subjectId", "group")])$group)
```

Inspecting the head of the table confirms the long format: each subject contributes \(K\) rows (one per category), and the imposed enrichment appears as larger `n` in `Condition1` for `Category0` only. The subject–group table should show six subjects in each of the three groups.

### Extract compositions and run permutation inference

The long table is aggregated to a `CategoryComposition` object: a subjects-by-categories count matrix with a named group factor. `ExtractClusterComposition()` performs that coercion and records provenance; missing subject–category combinations are zero-filled.

`CompareGroupCompositions()` accepts either a `CategoryComposition` or a long table directly. Here we extract explicitly so the subject-scale object is visible before inference.

```{r compare}
composition <- ExtractClusterComposition(long_table)
composition

perm_result <- CompareGroupCompositions(
  composition,
  method = "permutation",
  nPermutations = 200,
  seed = 42
)

perm_result
perm_result$omnibus
perm_result$contrasts
```

The printed `HellingerEnrichmentResult` summarizes the omnibus ratio and p-value. `perm_result$omnibus` holds the global effect size and unadjusted p-value; `perm_result$contrasts` holds all three pairwise contrasts with Holm-adjusted `pAdj` columns. Pairwise `contrastId` values follow sorted group-level order (`Condition1_vs_Control`, `Condition1_vs_Condition2`, `Condition2_vs_Control`). For the imposed enrichment, we expect `omnibus$effectSize > 1` and a small omnibus p-value, with `Condition1_vs_Control` (and often `Condition1_vs_Condition2`) carrying the enrichment, while `Condition2_vs_Control` remains near exchangeability (\(R \approx 1\), large p-values).

### Custom contrasts (and optional collapse)

Default output already includes all pairwise contrasts. Custom entries in `contrasts=` add named specifications on top of that default set: a length-2 character vector selects a subset pairwise comparison, while a list with `groups` and/or `collapse` remaps or restricts group levels before distances are computed. With three groups, a natural custom specification names the two scientifically primary comparisons—enriched versus Control, and the unmodified group versus Control—under stable identifiers for downstream tables.

```{r custom-contrasts}
perm_custom <- CompareGroupCompositions(
  composition,
  method = "permutation",
  nPermutations = 200,
  contrasts = list(
    condition1_vs_control = c("Control", "Condition1"),
    condition2_vs_control = c("Control", "Condition2")
  ),
  seed = 42
)

perm_custom$contrasts
```

The named contrasts reproduce those pairwise comparisons under user-chosen identifiers—useful when contrast tables are merged across analyses. Collapse becomes essential when experimental groups are relabeled for analysis (for example, merging dose cohorts into a single treated group) without rebuilding the count matrix by hand; see `?CompareGroupCompositions` for the `collapse` argument.

### Bayesian approach

The Bayesian method reuses the same contrast specifications but replaces fixed compositions with Dirichlet posterior draws. For vignette execution time, keep `nPosterior` and `nPermutations` small; production analyses should raise both so posterior means and intervals stabilize.

```{r bayes}
bayes_result <- CompareGroupCompositions(
  composition,
  method = "bayes",
  nPosterior = 20,
  nPermutations = 50,
  seed = 42
)

bayes_result$omnibus
bayes_result$contrasts
```

Interpretation differs slightly from permutation output. `bayes_result$omnibus$effectSize` is the posterior mean of \(R\), not the single observed ratio at the softened point estimate. The interval `effectCiLow`–`effectCiHigh` communicates whether \(R > 1\) remains stable under composition uncertainty. When the credible interval excludes 1, enrichment is supported even if the posterior-mean p-value is moderated by label shuffling within each draw. When the interval spans 1 but permutation p-values are small, sparse counts may be driving a fixed-composition signal that the Bayesian procedure correctly weakens.

### Plot contrasts

`PlotCompositionContrasts()` visualizes contrast-level effect sizes. A dotted vertical line at \(R = 1\) marks the exchangeability reference under this metric. When `showOmnibus = TRUE` (default), the omnibus posterior-mean or observed ratio appears as a dashed reference with its p-value annotated—contrasts can then be read against both unity and the global summary.

```{r plot-perm, fig.cap="Permutation contrast panel with omnibus reference."}
PlotCompositionContrasts(perm_result)
```

For permutation results, points are colored by whether Holm-adjusted `pAdj < 0.05`. For Bayes results, half-eye plots show the posterior density of each contrast's \(R\) with nested intervals.

```{r plot-bayes, fig.cap="Bayesian half-eye posterior densities for pairwise contrasts."}
PlotCompositionContrasts(bayes_result)
```

The two panels therefore display the same ratio scale with method-appropriate uncertainty. Permutation emphasizes discrete significance against the label null; Bayes overlays composition uncertainty on that same null via half-eye densities.

### Closing interpretation

On the synthetic cohort, both approaches should recover the imposed `Condition1` enrichment in `Category0`: omnibus and `Condition1_vs_Control` ratios above 1, with permutation p-values and Bayesian intervals consistent with a real between-group shift rather than label noise. The `Condition2_vs_Control` contrast should remain near \(R \approx 1\), illustrating that a significant omnibus result does not imply every pairwise group differs from Control. That pattern is expected here because the enrichment factor is large relative to multinomial sampling noise, Condition2 was never modified, and every subject contributes moderately dense counts across five categories.

The result does not generalize to all real cohorts. Sparse categories, unequal cell yields, or violated exchangeability (for example, batch confounded with group) can produce ratios that look enriched under fixed compositions but fail to hold once Dirichlet uncertainty is propagated. In those settings, the Bayesian approach is the more conservative assessment of whether an apparent contrast survives sampling variation in the compositions themselves.

Two practical constraints bound any analysis. First, each group must include at least two subjects, because within-group distances are defined from pairwise comparisons. Second, each subject must map to exactly one experimental group in the metadata; conflicting group labels for the same subject are rejected at composition build time. Within those constraints, the workflow—long counts, subject composition, contrast specification, ratio-based inference, and contrast plotting—extends from this synthetic example to real `ExtractClusterComposition()` inputs without changing the inferential definition of \(R\).
